---
title: "Reproducible Research - Programming Assignment 1"
output:
  html_document: default
---

### Introduction

Monitoring devices are being used by health cautious people as well as certain patients to record and track their daily physical activities. Translated to the number of steps taken, these data can help individuals to make adjustment to their daily workouts and even changing their life style for a healthier life.  

Using the data set produced from a device worn by a volunteer for 24 hours a day over two months we would like to analyze the measured data (number of steps), collected every 5 minutes to assess if there is a trend in his/her daily activities.

### Data Set

The data set can be downloaded from the site: 
[Dataset: Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) 

The variables included in this data set are:

* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)    
* date: The date on which the measurement was taken in YYYY-MM-DD format    
* interval: Identifier for the 5-minute interval in which measurement was taken    

The data set is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations (collection interval) in this data set. There are 288 (=24x12) collection intervals per day. The subject was wearing the device for 61 consecutive days between 10/1/2012-11/30/2012, which comes to a total of 17,568 (288x61) observations. 

### Analysis

**Loading and preprocessing the data**  

* Load the data set and the required libraries.  
* Remove NAs, and format the data.

```{r}
library(ggplot2)
library(plyr)
library(gcookbook) 
data <- read.csv2("repdata_data_activity/activity.csv",header=TRUE,sep=",",colClasses = "character")
data$steps <- as.numeric(data$steps)
data$interval <- as.numeric(data$interval)
data1 <- data[!is.na(data$steps),]
```

__*Total number of steps taken per day*__    

* Make a histogram of the total number of steps taken each day
* Calculate and report the mean and median of total number of steps taken per day
* Is there a patter in the plot that could indicate a trend in subject's activities. 

```{r}
#png(file = "plot14.png")
dailyTotal <- ddply(data1, "date", summarise, daily_total = sum(steps))
ggplot(dailyTotal, aes(x=date, y=daily_total)) +
    geom_bar(stat="identity", fill="lightblue", colr="black") +
    ggtitle("Total number of steps taken each day") +
    theme(axis.text.x = element_text(angle=90))

mean(dailyTotal$daily_total)
median(arrange(dailyTotal,daily_total)$daily_total)
#dev.off()
```

There is inconsistency in daily activities, since total number of steps is not uniform across all days. Few days have much smaller values than the visual average. To check if these values fall in a specific day of the week I calculated and plotted the average of the totals for days of the week, which shows the average of weekdays across the eight weeks of monitoring is much lower on Wednesday's. This could be because data for few Wednesdays are missing. More on this is discussed in section "Missing observations".

```{r}
dailyTotal$wkd <- as.factor(weekdays(as.POSIXlt(dailyTotal$date, format="%Y-%M-%d")))
weekly <- ddply(dailyTotal, "wkd", summarise, wkd_mean = mean(daily_total))
ggplot(weekly, aes(x=wkd, y=wkd_mean)) +
    geom_bar(stat="identity", fill="lightblue", colr="black") +
    xlab("Week Days") + ylab("Average of daily total") +
    ggtitle("Average of total number of steps grouped by week days")
```

__*Daily activity pattern*__   

* What is the average daily activity pattern?
    + Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days  
    + Which 5-minute interval, on average across all the days in the data set, contains the maximum number of steps?  

```{r}
intMean <- ddply(data1, "interval", summarise, Average_Interval = mean(steps))
ggplot(intMean,aes(x=c(1:nrow(intMean)),y=Average_Interval)) + geom_line() + geom_point() +
    xlab("Intervals (288 per day)") + ylab("Average number of steps") +
    ggtitle("Average number of steps taken per interval, across all (61) days")
maxval <- intMean[intMean$Average_Interval==max(intMean$Average_Interval),]
peak <- round(maxval[2],1)
interval <- rownames(maxval)
```

There is a clear pattern in daily activities. In early morning subject seems to be  moving around the house, occasionally. The activities pick up in the morning and fluctuates throughout the day with the maximum peak value of `r peak`, which occurs in interval `r interval`, around `r round(as.numeric(interval)/12,0)` AM.        

__*Imputing missing values*__  

The presence of missing days may introduce bias into some calculations or summaries of the data.  

* Calculate and report the total number of missing values in the data set (i.e. the total number of rows with NAs)
* Devise a strategy for filling in all of the missing values in the data set. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.  
* Create a new data set that is equal to the original data set but with the missing data filled in.  
* Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
    + Do these values differ from the estimates from the first part of the assignment? 
    + What is the impact of imputing missing data on the estimates of the total daily number of steps?  

__*Missing observations*__  

```{r}
miss_obs <- subset(data, is.na(data$steps), drop=F)
uniq_missing_date <- unique(miss_obs$date)
wday <- as.factor(weekdays(as.POSIXlt(uniq_missing_date, format="%Y-%M-%d")))
uniq_missing_date
wday
```

The original data set has `r nrow(miss_obs)` NAs in the steps variable, which belong to `r length(uniq_missing_date)` days. That is, there isn't a single observation for either of the missing days, perhaps the subject was not wearing the device on those days.   

__*Strategy to compensate for missing observations*__  

Estimating the mean or median for the missing days is not possible due to lack of measurement. For total daily steps we could use the data in "weekly" bar graph (calculated above) for any missing day of the week, but these data can not be used for replacing the missing interval data. However, we could use the interval averages for calculating both daily total and average interval.   

To create the new data set we merge the original data with the interval mean (intMean) data frame to match the intervals and to bring the associated data (Average_Interval) into one table. Then, by walking through the new table replace the missing steps (NA) with the average of corresponding interval.   

```{r}
newdata <- merge(data,intMean,by="interval")
for (i in 1:nrow(newdata)) if (is.na(newdata$steps[i])) newdata$steps[i] <- newdata$Average_Interval[i]    
```

Plot of new data set for daily total: 

```{r}
# New Data set: Plot total steps per day
dailyTotNA <- ddply(newdata, "date", summarise, daily_total_noNA = sum(steps))
ggplot(dailyTotNA, aes(x=date, y=daily_total_noNA)) + 
    geom_bar(stat="identity", fill="lightblue", colr="black") +
    theme(axis.text.x = element_text(angle=90)) +
    xlab("Date") + ylab("Daily Total Number of Steps") +
    ggtitle("New data: Total number of steps taken per day, across all (288) intervals")

mean(dailyTotNA$daily_total_noNA)
median(arrange(dailyTotNA,daily_total_noNA)$daily_total_noNA)
```

As expected, there are no significant differences between the medians and means of old and new data. 

Plot of new data set for interval means:

```{r}
# New Data: Plot average of steps for each interval, accross all days
intMeanNA <- ddply(newdata, "interval", summarise, Average_noNA = mean(steps))
ggplot(intMeanNA,aes(x=c(1:nrow(intMeanNA)),y=Average_noNA)) + geom_line() + geom_point() +
    xlab("Intervals (288 per day)") + ylab("Average number of steps") +
    ggtitle("New data: Average number of steps taken per interval, across all (61) days")
```

As expected, because we used the average to estimate the missing data, there isn't a noticeable change between patterns of data with or without replacing NAs. 

__*Weekdays vs Weekends*__  

Are there differences in activity patterns between weekdays and weekends?

```{r}
# Create a factor and convert day of the week to weekday and weekend.

newdata$wkd <- as.factor(weekdays(as.POSIXlt(newdata$date, format="%Y-%M-%d")))
levels(newdata$wkd)[c(3:4)] <- "Weekend"
levels(newdata$wkd)[c(1,2,4,5,6)] <- "Weekday"

wkdayend <- ddply(newdata, c("wkd", "interval"), summarise, Int_mean = mean(steps))
wkdayend$INTERVAL <- rep(1:288,2)
ggplot(wkdayend,aes(x=INTERVAL, y=Int_mean, color=wkd)) + geom_line(size=1) + 
    geom_point(size=2, shape=21, fill="white") + facet_grid(wkd ~ .) + 
    xlab("Activity Interval (5 mins)") + ylab("Mean of total steps per interval") + 
    ggtitle("Activity Recorded Over Weekdays and Weekends") +
    theme(legend.position="none", strip.text=element_text(face="bold", size=rel(1)),
          strip.background=element_rect(fill="lightblue", colour="black", size=1))
```

On the average, activities are the same during weekdays and weekends. This could very well indicate that activities on weekends has to be more intensive. This can be verified in a boxplot, shown below. To get a better picture, I removed the noise (data between evening and early morning) from the data set.  

```{r}
highActive <- wkdayend[wkdayend$INTERVAL>70 & wkdayend$INTERVAL<260,]
ggplot(highActive, aes(x=wkd, y=Int_mean)) + geom_boxplot()
```